---
title: "BGGN-213: Visualization with R Lecture #8 (Part 1)"
output: github_document
date: 2017-10-23   (14:04:51 PDT on Mon, Oct 23)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Circle points function
Lets write a wee function to create some points on the circumference of a circle. We will use the equation of a circle with origin (j, k) and radius r: `x(t) = r cos(t) + j`  and `y(t) = r sin(t) + k` See for example: http://www.math.com/tables/geometry/circles.htm
      

```{r}
cpoints <- function(radius=5, center=c(0,0), angle=seq(0,360,length=15)) {
  return( cbind( x=radius*cos(angle)+center[1],
                 y=radius*sin(angle)+center[2] ))
}
```

Lets print out p and make a quick plot
```{r}
p <- cpoints()
p
plot( p )
```

Calculate some summary statistics for x and y. Note here we use the **apply()** function to call the **summary()** function on the cols by setting the `margin=2` argument. What would setting `margin=1 return?

```{r}
apply(p, 2, summary)
```


## Tidy the figure
Lets make the plot square and increase axis font sizes, change ploting character and increase its size. 

```{r}
par(pty="s", cex=1.3)
plot(p, xlim=c(-6,6), ylim=c(-6,6), xlab="x", ylab="y", pch=16, cex=2)
```

For my lecture slides I will make a PDF and have the points and font in white with transparent background (the default when we save a PDF from code rather than the RStudio `Export` button).

```{r}
pdf(file="Rplot.pdf")
par(pty="s", cex=1.3)
plot(p, xlim=c(-6,6), ylim=c(-6,6), xlab="x", ylab="y", col="white", col.axis="white", col.lab="white", pch=16, cex=2)
dev.off()
```



## Non optimal plot in this instance
But showing it as a time-series obscures the relationship. This is still better than a simple table.

```{r}
plot(p[,1], typ="b", col="green", lwd=3)
points(p[,2], typ="b", col="blue", lwd=3)
```



Or we can use ggplot2 but note that ggplot2 needs a **data.frame** object as input!

```{r}
library(ggplot2)
pc <- as.data.frame(p)
ggplot(pc, aes(x,y)) + geom_point() + coord_equal()
```

For some other plots in this lecture we will use the more complicated but more versitile **ggplot2** package. We will cover the basics of ggplot later.

## Side-Note: GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Session Information
```{r}
sessionInfo()
```

